The Talking Heads experiment

### 2015-04-29 chap2/chap2.tex
part 1 of martin haspelmath's comments

Let me make a distinction between the physical aspects of 
a cognitive agent and the mental aspects. The physical aspects include the 
agent's body, the sensors and articulators, 
the physical location, the objects in this location, and the other
agents physically present in the environment. The 
mental aspects include the agent's repertoire of behaviours, 
the brain structures and processes
performing categorisations of reality, his
memory, lexicon, grammar, and so on. 
In the case of humans, these two aspects are intimately 
connected and indivisible. We cannot teleport our mental 
faculties into another body, or into another copy of 
our body, even though there have been speculations that 
we could in the future record human 
brain states,\footnote{Such visions of the future have been put forward
by \cite{Moravec:1995}. Neither current artificial intelligence technology nor
the state of the art in brain state recording are anywhere 
near to realising these visions.}
I believe that this will still not enable teleportation
because in human brains there is no distinction between 
hardware and software. The architecture of a human brain, 
the physical connections between cells, and the 
biochemical processes in each cell determine the brain's behaviour.
There is no separation between a {\itshape brain program} and
an interpreter that reads brain programs and 
executes them. The brain is a special-purpose 
hardware device which is unique to each individual. 
To copy such a device we would have to rebuild it physically,
atom by atom, and integrate it in an exact copy of the 
same body.

@@

However, in the case of computer-based \is{teleportation}
artificial agents, we {\itshape can} 
make the distinction. It is possible to capture the mental state
of an agent in software, load this into a physical 
body, and then operate 
the agent. Afterwards the agent can extract himself
again from the body, teleport himself 
to another physical location through a data transmission network
like the Internet, get instantiated there in another
body, and experience another reality and physically meet
other agents. This is exactly how we have implemented the 
Talking Heads experiment.\footnote{The agent teleportation infrastructure is in itself a 
fascinating non-trivial engineering project. Contributions 
from Angus McIntyre, Alexis Agahi, Sylvere Tajan and
Frederic Kaplan are gratefully acknowledged (\citealt{McIntyre:1999}). 
The fact that agents can teleport proves that we are dealing 
with a truly distributed multi-agent system. It also 
introduces physical parallelism in the agent-agent and 
agent-environment interactions. 
For a general introduction into multi-agent system
technologies and design methodologies, 
see \cite{Ferber:1999}.}
There are on the one hand
the physical structures, which I will refer to as the {\it
robot bodies}. They are installed in different physical 
locations somewhere in the world and connected
with each other through the Internet. Then there is 
a population of software structures that 
are occasionally loaded and instantiated in specific
robots. I will call these software structures
{\itshape virtual agents}. 
A {\itshape real agent} (a Talking Head) 
only exists when the virtual agent is
loaded in a physical robot body.

@@

The Guessing Game is played between two 
physically instantiated agents. Agents in 
a virtual state must queue up to have access to one of the 
robot bodies installed in a particular site before
they can play the game. One agent 
plays the role of {\itshape speaker} and the other then plays the 
role of {\itshape hearer}. Agents take turn playing games 
so that all of them develop the capacity to be speaker or hearer.
A human experimenter can pick one of these roles and 
play the game instead of an artificial agent.

The speaker first looks at one area of the white board and 
directs the attention of the hearer to the same area.\footnote{I will explain later how exactly agents
decide on a particular scene and how they are able to 
draw each other's attention to specific areas of the 
white board.}
The objects located in this area constitute the {\itshape context}. 
The speaker then chooses one object from the context, which 
I will call the {\itshape topic}, and gives a verbal hint
to the hearer. The {\itshape verbal hint} is an expression
that identifies the topic 
with respect to the other objects in the context. For example, 
if the context contains a red square, a blue triangle, 
and a green circle, then the speaker may say something like 
\smplenquote{the red one} to identify the red square as the topic. 
If the context contains also a red triangle, he has to be more 
precise and say something like \smplenquote{the red square} to delineate
the topic from the red triangle as well as
from the blue square. 
Of course, the Talking Heads do not say \smplenquote{the red square}
but use their own language and concepts which are not 
necessarily the same as those used in English. For example, 
they may say \smplenquote{malewina} to mean [\textsc{upper extreme}-\textsc{left low}-\textsc{redness}].

@@

The Guessing Game is one of the common things we do with language. 
For example, I play a similar game when I sit with a friend
at the dining table and say \smplenquote{could you give me the salt}. 
If she guesses correctly what I mean and hands 
me the salt, the game has succeeded. If she looks at me 
with a puzzled face (maybe she does not speak English) or hands me
the salmon instead of the salt, the game has failed. In that case, 
I can gesture in the direction of the salt and say `no, no, the {\itshape salt}
please', and then she hopefully realises and gives the salt to me. Failure 
is common in natural language dialogues and may be caused by 
many factors. For example, the salmon could indeed 
be close to the salt and my pronunciation of the word \smplenquote{salt}
may have sounded a bit like \smplenquote{salmon}, perhaps because there was
loud music playing in the background or perhaps
my friend does not understand English. 
A failure is often 
an opportunity to negotiate how something will be expressed
in the future. For example, the hearer may pick up a new word
or the speaker may realise that a certain word is 
not appropriate in this particular context.

@@

The environment of the Talking Heads is 
not fixed. The human experimenter 
may change the position of objects, add new kinds of 
objects, or eliminate
others. Consequently a strategy of naming individual objects
will not work. It would lead to a proliferation of 
proper names and it would require the Talking Heads to
recognise objects, which is very difficult to do.\footnote{For a thorough exposition of the difficulties of 
object recognition, see \cite{Ullman:1996}. 
Object constancy comes fairly late in the 
acquisition of a child's ontology, as Piaget's conservation
experiments have shown. Language probably plays an important
role in forming the notion of an object.}
Indeed, humans don't exclusively use proper names in 
natural language conversations 
either. We say \smplenquote{could you give me the red small square} as 
opposed to \smplenquote{could you give me O\_143}. Natural language
words like \smplenquote{red} or \smplenquote{small} name perceptually grounded categories
and syntactic structures indicate how they should
be combined and used to find the topic. The relation between 
a language expression and its referent is therefore 
always indirect. This is summarised
in the semiotic square (\figref{triangle}), which I will use throughout this book to 
help understand and analyse the nature of language communication. \is{semiotic square}
The semiotic square relates the four entities
involved in a verbal interaction: 
\begin{itemize}
\item An {\itshape utterance}, such as \smplenquote{small red square}, which is 
transmitted as a physical 
signal from one agent to another one through the 
external environment. It is written between double quotes. 
\item A {\itshape meaning}, which consists of categories like [\textsc{red}], [\textsc{small}], 
or combinations of categories, like \{[\textsc{red}] [\textsc{small}]\}. Labels of 
categories are written in capital letters between 
square brackets. 
\clearpage
\item An {\itshape image segment}, denoted as \emph{S143}, which is 
a segment of the image perceived through the camera.
\item A {\itshape referent}, like object \emph{O143}, which is 
an entity in the real world. 
\end{itemize}

@@

I will use the following terms for denoting the 
processes speakers and hearers go through while traversing
the relations in the semiotic square as they play a language
game (\figref{triangle2a}). 
A similar framework, but emphasising the 
language production side, has been described in great
detail by \cite{Levelt:1989}. This book also provides a wealth
of psychological evidence that these processes must 
be going on and expands the phonetic and phonological 
side. An example of a detailed architecture inspired by 
generative grammar is discussed in: \cite{Jackendoff:1997}. 
Generative approaches to language attempt to define a language by 
generating its set of possible utterances. Interpretations
are constructed from the syntactic structure derived by 
the generative grammar. In this book, we are interested 
in the mapping from communicative intent 
in a perceived reality to an utterance and back. The knowledge
and skill needed to solve this problem is different from 
that need to systematically generate the set of sentences in a 
language and their possible interpretations.
\begin{enumerate}
\item The speaker as well as the hearer {\itshape perceive} reality
by capturing an image through the camera, segmenting 
the image into coherent units, and deriving various 
sensory characteristics about each image segment, 
such as the colour, size, movement.

@@

\item The speaker must then {\itshape conceptualise} the 
scene on the basis of this perception. He must find a set of
categories or a conjunctive combination of 
categories that distinguishes the referent from the other objects in 
the context, and which will act as the meaning of 
his communication. For example, he might choose [\textsc{blue}] if the
topic has a blue colour and all the other objects in the 
context are not blue.

\item The speaker must then {\itshape verbalise} this conceptualisation. 
He must use his language system to find words and 
syntactic constructions expressing this meaning.
For example, he might choose \smplenquote{blauw} (if he speaks
Dutch), or \smplenquote{bleu} (if he speaks French),
to convey the category [\textsc{blue}].

\item The hearer must engage in similar tasks but now
going in the reverse direction. He must
{\itshape interpret} the utterance to find out which
conceptualisation constitutes the meaning.

\item Then the hearer must {\itshape apply} this meaning
to see what referent was intended. The hearer has
also perceived the scene in terms of a set of 
segments and now uses 
the meaning to identify the segment that
could have been the one intended by the speaker.

\item Finally the hearer {\itshape acts} upon the outcome of meaning
application. He points to the topic he has identified. This is the 
step where both agents co-ordinate their 
behaviour through the external world. 
\end{enumerate}

@@

Each of these activities requires knowledge and/or skill
(summarised in the table below). Perceiving requires 
visual processes capable of segmenting images and deriving image 
segments. Conceptualisation requires an ontology, 
a repertoire of perceptually grounded distinctions
that can be applied to a segmented image to yield 
distinctive categories or category combinations that 
may constitute the meaning of the utterance. Verbalising 
a conceptualisation requires 
a {\itshape lexicon} that maps parts of the meaning to words and 
a {\itshape syntax} that specifies how to organise individual words
into a larger complex. 
The hearer must use similar 
knowledge sources in the other direction. He must use 
his lexicon and syntax to reconstruct the meanings expressed
by the utterance, and then use the 
ontology again to apply the meaning to the present 
context to find the referent (see \tabref{tab:2:findthereferent}).

@@

I deliberately left an important aspect of language out. 
In the case of physical agents, the form cannot be transmitted
directly but needs to be articulated in speech sounds, 
written signs, or gestures, to create a true utterance. This additional 
complexity will not be discussed further in this book -- even though 
it is a fascinating topic in its own right.\footnote{See for a state of the art review: \cite{Clark:1990}. 
We have already been conducting quite extensive 
research in our group on the origin of sound systems 
using similar principles as the ones discussed in this 
book for the origins of word semantics. 
See  \cite{DeBoer:1997}. The complex 
adaptive systems approach underlying this phonetics
work was already foreshadowed in work of phoneticians
like \cite{Liljencrants:1972}.}
To make a verbal interaction nevertheless complete, agents
are given a repertoire of consonants and vowels with 
which they can make random syllables and syllable combinations, 
like \smplenquote{wabido}, \smplenquote{bimaku}, etc. The articulation and 
recognition of these syllables is assumed to be acquired
already and transmission is engineered to be error-free. 
This way, our attention can be focused on how
ontologies, lexicons, and grammars may emerge.

@@

Every physical Talking Heads set-up features a {\itshape waiting room} in 
which agents are stored, ready to be loaded into the 
robotic infrastructure or to be teleported to another
physical site. A game starts when two agents are chosen
at random from this waiting room and loaded into 
the two physical robots. The
internal architecture of each agent is connected with 
the sensori-motor apparatus of the robot so that they can 
receive the sensory data streams recorded by the camera
and control the movements of the robots. Then the general 
control system assigns randomly the role of speaker and hearer
to each of the agents and the game can begin.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap2/figs/hpos.pdf}}
\caption{Three examples of segmented images (from top to bottom). Left and right images 
show the perceived and segmented images for two agents respectively. These images are always
different because they are dependent on the position of the agent. 
The topic is indicated by a dashed bounding box in the 
image of the speaker (on the right for the first case and on the left for the others). 
Segments which are too small are ignored. The topics have all been conceptualised
as being to the right and so the same word 
\smplenquote{gofubo} has been used to successfully refer to them.}
\label{f:plate10}
\end{figure}

@@

Next low-level visual processes gather information about each \is{sensory channels}
segment, such as its average colour, size, shape,
the position with respect to horizontal or vertical axes. 
Each process outputs its information on a {\itshape sensory 
channel} scaled between 0.0 and 1.0. The mechanisms from 
the conceptualisation layer that subsequently use this information
can operate on any kind of sensory channel. 
I will assume for the rest of this chapter that the low-level
routines produce each only three types of information, sent
on the following sensory channels: 
\begin{itemize}
\item The first channel is called \textsc{hpos} (horizontal position). 
It contains the x-midpos\-ition of a segmented object within
the field of view of the robot. 
\item The second channel is called \textsc{vpos} (vertical position). It
specifies the y-midpos\-ition of the segmented object. 
\item The third channel is called \textsc{gray} and contains 
the average gray-scale of the object. 
\end{itemize}
Later on additional channels will be introduced.

@@

Children acquire their first words around the first year
of life.\footnote{Representative work in the study of child language
learning focuses mostly on the acquisition of specific meanings.
See  \cite{Gleitman:1994}, 
\cite{Clark:1993}, \cite{Bowerman:1996}.}
Most people believe that they do this 
as a result of hearing a particular word repeated several times in a
certain context and gradually abstracting an
association between a word and a meaning. But 
how is it that they know which meaning to associate with a 
particular word? How is it that word acquisition goes
so rapid? We will follow a different 
approach, which leaves an open question whether
this applies to human word acquisition as well. The
approach will be selectionist. The agents {\itshape construct} 
hypotheses, either on the basis of one specific case 
where they guess through a non-verbal strategy what 
the meaning of an unknown word might be, 
or they have simply invented a new word because
they do not have one yet. The hypotheses are 
then tried out in subsequent games and either receive confirmation
or are falsified. As a side effect of this local behaviour, 
a global self-organising dynamic process arises leading gradually to coherence.

@@

Here is the simplest possible instance of a
language game, based on the scene in \figref{scene1-1}. 
The speaker, {\bfshape a1}, has picked the triangle as the topic.
To give a verbal hint, he needs 
to conceptualise this topic, which means in this 
specific case, to find a 
category (or set of categories) which distinguishes the 
topic from the other objects in the context. Here 
the context contains only one additional 
object, namely the rectangle. The category
[\textsc{vpos} 0.0–0.5]$_{\mathbf a1}$ (lower) fits the criteria. 
[\textsc{vpos} 0.0–0.5] is valid when \textsc{vpos}$ < 0.5$
which is the case for the triangle but not for the rectangle. 
{\bfshape a1} has an association in his lexicon 
relating [\textsc{vpos} 0.0–0.5]$_{\mathbf  a1}$ with the word \smplenquote{lu}, 
{\bfshape a1} retrieves this word and transmits it to 
the hearer, which is agent {\bfshape a2}.

{\bfshape a2} has stored in his lexicon an
association between \smplenquote{lu} and [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$, and so
hypothesises that [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$ must be the meaning of \smplenquote{lu}. 
When this category is applied to the present scene, in other words,
when {\bfshape a2} filters out the objects whose value for \textsc{vpos} do not 
fall in the region $[0.0–0.5]$, only one 
remaining object is obtained, the triangle. Hence {\bfshape a2} concludes that 
this must be the topic and points to it.

@@

There are (at least) two ways in which a game similar to game 125, 
but played earlier might have failed. First of all it can be that
the speaker does not have a word yet for the meaning he wants
to convey. The speaker then invokes a strategy 
to repair this failure. The simplest strategy is to 
create a new word for the present meaning. This is how 
{\bfshape a1} might have created the word \smplenquote{lu}, and associated it 
with [\textsc{vpos} 0.0–0.5] in his lexicon. Such simple constructive steps cause
new words to enter the lexicon.

Second, it can be that the hearer does not know the word. The 
game then fails and the speaker points to 
the topic so that the hearer can make 
an educated guess what the meaning might have been:
If the hearer is able to recover a possible topic
from the non-verbal hint given by the speaker, 
he himself can seek a distinctive category or set of categories 
that delineates this topic from the other objects in 
the context just as the speaker has done. It is 
possible (although not necessary as we will see)
that the hearer {\bfshape a2} arrives at his 
own version of the same category, namely [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$. The 
hearer now stores in his lexicon a new association between the 
form heard, \smplenquote{lu}, and the guessed meaning [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$. 
Based on this extended lexicon, he will succeed in the same game 
in the future and he can use \smplenquote{lu} himself to verbalise [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$
when talking to other agents. This is the way new words
spread in the population.

@@

The observant reader will have noticed immediately that 
I have swept some important difficulties under the carpet. 
First, it could very well be that unknown to the speaker 
another agent had already used the word \smplenquote{lu} 
for another meaning, so that there are now 
two alternative meanings for \smplenquote{lu} in the lexicon. \smplenquote{lu} 
has become ambiguous. Second, it may be the hearer {\bfshape a2} already 
had a word for [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$, for example \smplenquote{bomida}, 
so that there are now two synonyms for [\textsc{vpos} 0.0–0.5] in the lexicon. 
Synonymy and ambiguity are very common in 
natural languages and unavoidable when a group of 
distributed autonomous agents creates a language without
a central co-ordinator. This implies that 
the agents' lexicon must be sophisticated enough 
to support multiple associations. An agent must be able to 
store the different meanings being used for the same word, and the 
different words being used for the same meaning.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap2/figs/competition-1.pdf}}
\caption{\label{competition1}The graph shows the competition between different forms for
expressing one meaning in a population of 20 agents. The graph plots 
the frequency of each form in the population, more precisely the percentage
of agents that prefer a particular word. A complex dynamic
process unfolds with periods where one word (first \smplenquote{xu} and then \smplenquote{fepi}) dominates.} 
\end{figure}

>>>>

Let me make a distinction between the physical aspects of 
a cognitive agent and the mental aspects. The physical aspects include the 
agent's body, the sensors and articulators, 
the physical location, the objects in this location, and the other
agents physically present in the environment. The 
mental aspects include the agent's repertoire of behaviours, 
the brain structures and processes
performing categorisations of reality, his
memory, lexicon, grammar, and so on. 
In the case of humans, these two aspects are intimately 
connected and indivisible. We cannot teleport our mental 
faculties into another body, or into another copy of 
our body, even though there have been speculations that 
we could in the future record human 
brain states,\footnote{Such visions of the future have been put forward
by \cite{Moravec:1995}. Neither current artificial intelligence technology nor
the state of the art in brain state recording are anywhere 
near to realising these visions.}
I believe that this will still not enable teleportation
because in human brains there is no distinction between 
hardware and software. The architecture of a human brain, 
the physical connections between cells, and the 
biochemical processes in each cell determine the brain's behaviour.
There is no separation between a {\scshape brain program} and
an interpreter that reads brain programs and 
executes them. The brain is a special-purpose 
hardware device which is unique to each individual. 
To copy such a device we would have to rebuild it physically,
atom by atom, and integrate it in an exact copy of the 
same body.

@@

However, in the case of computer-based \is{teleportation}
artificial agents, we {\itshape can} 
make the distinction. It is possible to capture the mental state
of an agent in software, load this into a physical 
body, and then operate 
the agent. Afterwards the agent can extract himself
again from the body, teleport himself 
to another physical location through a data transmission network
like the Internet, get instantiated there in another
body, and experience another reality and physically meet
other agents. This is exactly how we have implemented the 
Talking Heads experiment.\footnote{The agent teleportation infrastructure is in itself a 
fascinating non-trivial engineering project. Contributions 
from Angus McIntyre, Alexis Agahi, Sylvere Tajan and
Frederic Kaplan are gratefully acknowledged (\citealt{McIntyre:1999}). 
The fact that agents can teleport proves that we are dealing 
with a truly distributed multi-agent system. It also 
introduces physical parallelism in the agent-agent and 
agent-environment interactions. 
For a general introduction into multi-agent system
technologies and design methodologies, 
see \cite{Ferber:1999}.}
There are on the one hand
the physical structures, which I will refer to as the {\scshape
robot bodies}. They are installed in different physical 
locations somewhere in the world and connected
with each other through the Internet. Then there is 
a population of software structures that 
are occasionally loaded and instantiated in specific
robots. I will call these software structures
{\scshape virtual agents}. 
A {\scshape real agent} (a Talking Head) 
only exists when the virtual agent is
loaded in a physical robot body.

@@

The Guessing Game is played between two 
physically instantiated agents. Agents in 
a virtual state must queue up to have access to one of the 
robot bodies installed in a particular site before
they can play the game. One agent 
plays the role of {\scshape speaker} and the other then plays the 
role of {\scshape hearer}. Agents take turn playing games 
so that all of them develop the capacity to be speaker or hearer.
A human experimenter can pick one of these roles and 
play the game instead of an artificial agent.

The speaker first looks at one area of the white board and 
directs the attention of the hearer to the same area.\footnote{I will explain later how exactly agents
decide on a particular scene and how they are able to 
draw each other's attention to specific areas of the 
white board.}
The objects located in this area constitute the {\scshape context}. 
The speaker then chooses one object from the context, which 
I will call the {\scshape topic}, and gives a verbal hint
to the hearer. The {\scshape verbal hint} is an expression
that identifies the topic 
with respect to the other objects in the context. For example, 
if the context contains a red square, a blue triangle, 
and a green circle, then the speaker may say something like 
\textit{the red one} to identify the red square as the topic. 
If the context contains also a red triangle, he has to be more 
precise and say something like \textit{the red square} to delineate
the topic from the red triangle as well as
from the blue square. 
Of course, the Talking Heads do not say \textit{the red square}
but use their own language and concepts which are not 
necessarily the same as those used in English. For example, 
they may say \textit{malewina} to mean [\textsc{upper extreme}-\textsc{left low}-\textsc{redness}].

@@

The Guessing Game is one of the common things we do with language. 
For example, I play a similar game when I sit with a friend
at the dining table and say \textit{could you give me the salt}. 
If she guesses correctly what I mean and hands 
me the salt, the game has succeeded. If she looks at me 
with a puzzled face (maybe she does not speak English) or hands me
the salmon instead of the salt, the game has failed. In that case, 
I can gesture in the direction of the salt and say \textit{no, no, the {\scshape salt}
please}, and then she hopefully realises and gives the salt to me. Failure 
is common in natural language dialogues and may be caused by 
many factors. For example, the salmon could indeed 
be close to the salt and my pronunciation of the word \textit{salt}
may have sounded a bit like \textit{salmon}, perhaps because there was
loud music playing in the background or perhaps
my friend does not understand English. 
A failure is often 
an opportunity to negotiate how something will be expressed
in the future. For example, the hearer may pick up a new word
or the speaker may realise that a certain word is 
not appropriate in this particular context.

@@

The environment of the Talking Heads is 
not fixed. The human experimenter 
may change the position of objects, add new kinds of 
objects, or eliminate
others. Consequently a strategy of naming individual objects
will not work. It would lead to a proliferation of 
proper names and it would require the Talking Heads to
recognise objects, which is very difficult to do.\footnote{For a thorough exposition of the difficulties of 
object recognition, see \cite{Ullman:1996}. 
Object constancy comes fairly late in the 
acquisition of a child's ontology, as Piaget's conservation
experiments have shown. Language probably plays an important
role in forming the notion of an object.}
Indeed, humans don't exclusively use proper names in 
natural language conversations 
either. We say \textit{could you give me the red small square} as 
opposed to \textit{could you give me O\_143}. Natural language
words like \textit{red} or \textit{small} name perceptually grounded categories
and syntactic structures indicate how they should
be combined and used to find the topic. The relation between 
a language expression and its referent is therefore 
always indirect. This is summarised
in the semiotic square (\figref{triangle}), which I will use throughout this book to 
help understand and analyse the nature of language communication. \is{semiotic square}
The semiotic square relates the four entities
involved in a verbal interaction: 
\begin{itemize}
\item An {\scshape utterance}, such as \textit{small red square}, which is 
transmitted as a physical 
signal from one agent to another one through the 
external environment. It is written between double quotes. 
\item A {\scshape meaning}, which consists of categories like [\textsc{red}], [\textsc{small}], 
or combinations of categories, like \{[\textsc{red}] [\textsc{small}]\}. Labels of 
categories are written in capital letters between 
square brackets. 
\clearpage
\item An {\scshape image segment}, denoted as \emph{S143}, which is 
a segment of the image perceived through the camera.
\item A {\scshape referent}, like object \emph{O143}, which is 
an entity in the real world. 
\end{itemize}

@@

I will use the following terms for denoting the 
processes speakers and hearers go through while traversing
the relations in the semiotic square as they play a language
game (\figref{triangle2a}). 
A similar framework, but emphasising the 
language production side, has been described in great
detail by \cite{Levelt:1989}. This book also provides a wealth
of psychological evidence that these processes must 
be going on and expands the phonetic and phonological 
side. An example of a detailed architecture inspired by 
generative grammar is discussed in: \cite{Jackendoff:1997}. 
Generative approaches to language attempt to define a language by 
generating its set of possible utterances. Interpretations
are constructed from the syntactic structure derived by 
the generative grammar. In this book, we are interested 
in the mapping from communicative intent 
in a perceived reality to an utterance and back. The knowledge
and skill needed to solve this problem is different from 
that need to systematically generate the set of sentences in a 
language and their possible interpretations.
\begin{enumerate}
\item The speaker as well as the hearer {\scshape perceive} reality
by capturing an image through the camera, segmenting 
the image into coherent units, and deriving various 
sensory characteristics about each image segment, 
such as the colour, size, movement.

@@

\item The speaker must then {\scshape conceptualise} the 
scene on the basis of this perception. He must find a set of
categories or a conjunctive combination of 
categories that distinguishes the referent from the other objects in 
the context, and which will act as the meaning of 
his communication. For example, he might choose [\textsc{blue}] if the
topic has a blue colour and all the other objects in the 
context are not blue.

\item The speaker must then {\scshape verbalise} this conceptualisation. 
He must use his language system to find words and 
syntactic constructions expressing this meaning.
For example, he might choose \textit{blauw} (if he speaks
Dutch), or \textit{bleu} (if he speaks French),
to convey the category [\textsc{blue}].

\item The hearer must engage in similar tasks but now
going in the reverse direction. He must
{\scshape interpret} the utterance to find out which
conceptualisation constitutes the meaning.

\item Then the hearer must {\scshape apply} this meaning
to see what referent was intended. The hearer has
also perceived the scene in terms of a set of 
segments and now uses 
the meaning to identify the segment that
could have been the one intended by the speaker.

\item Finally the hearer {\scshape acts} upon the outcome of meaning
application. He points to the topic he has identified. This is the 
step where both agents co-ordinate their 
behaviour through the external world. 
\end{enumerate}

@@

Each of these activities requires knowledge and/or skill
(summarised in the table below). Perceiving requires 
visual processes capable of segmenting images and deriving image 
segments. Conceptualisation requires an ontology, 
a repertoire of perceptually grounded distinctions
that can be applied to a segmented image to yield 
distinctive categories or category combinations that 
may constitute the meaning of the utterance. Verbalising 
a conceptualisation requires 
a {\scshape lexicon} that maps parts of the meaning to words and 
a {\scshape syntax} that specifies how to organise individual words
into a larger complex. 
The hearer must use similar 
knowledge sources in the other direction. He must use 
his lexicon and syntax to reconstruct the meanings expressed
by the utterance, and then use the 
ontology again to apply the meaning to the present 
context to find the referent (see \tabref{tab:2:findthereferent}).

@@

I deliberately left an important aspect of language out. 
In the case of physical agents, the form cannot be transmitted
directly but needs to be articulated in speech sounds, 
written signs, or gestures, to create a true utterance. This additional 
complexity will not be discussed further in this book -- even though 
it is a fascinating topic in its own right.\footnote{See for a state of the art review: \cite{Clark:1990}. 
We have already been conducting quite extensive 
research in our group on the origin of sound systems 
using similar principles as the ones discussed in this 
book for the origins of word semantics. 
See  \cite{DeBoer:1997}. The complex 
adaptive systems approach underlying this phonetics
work was already foreshadowed in work of phoneticians
like \cite{Liljencrants:1972}.}
To make a verbal interaction nevertheless complete, agents
are given a repertoire of consonants and vowels with 
which they can make random syllables and syllable combinations, 
like \textit{wabido}, \textit{bimaku}, etc. The articulation and 
recognition of these syllables is assumed to be acquired
already and transmission is engineered to be error-free. 
This way, our attention can be focused on how
ontologies, lexicons, and grammars may emerge.

@@

Every physical Talking Heads set-up features a {\scshape waiting room} in 
which agents are stored, ready to be loaded into the 
robotic infrastructure or to be teleported to another
physical site. A game starts when two agents are chosen
at random from this waiting room and loaded into 
the two physical robots. The
internal architecture of each agent is connected with 
the sensori-motor apparatus of the robot so that they can 
receive the sensory data streams recorded by the camera
and control the movements of the robots. Then the general 
control system assigns randomly the role of speaker and hearer
to each of the agents and the game can begin.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap2/figs/hpos.pdf}}
\caption{Three examples of segmented images (from top to bottom). Left and right images 
show the perceived and segmented images for two agents respectively. These images are always
different because they are dependent on the position of the agent. 
The topic is indicated by a dashed bounding box in the 
image of the speaker (on the right for the first case and on the left for the others). 
Segments which are too small are ignored. The topics have all been conceptualised
as being to the right and so the same word 
\textit{gofubo} has been used to successfully refer to them.}
\label{f:plate10}
\end{figure}

@@

Next low-level visual processes gather information about each \is{sensory channels}
segment, such as its average colour, size, shape,
the position with respect to horizontal or vertical axes. 
Each process outputs its information on a {\scshape sensory 
channel} scaled between 0.0 and 1.0. The mechanisms from 
the conceptualisation layer that subsequently use this information
can operate on any kind of sensory channel. 
I will assume for the rest of this chapter that the low-level
routines produce each only three types of information, sent
on the following sensory channels: 
\begin{itemize}
\item The first channel is called \textsc{hpos} (horizontal position). 
It contains the x-midpos\-ition of a segmented object within
the field of view of the robot. 
\item The second channel is called \textsc{vpos} (vertical position). It
specifies the y-midpos\-ition of the segmented object. 
\item The third channel is called \textsc{gray} and contains 
the average gray-scale of the object. 
\end{itemize}
Later on additional channels will be introduced.

@@

Children acquire their first words around the first year
of life.\footnote{Representative work in the study of child language
learning focuses mostly on the acquisition of specific meanings.
See  \cite{Gleitman:1994}, 
\cite{Clark:1993}, \cite{Bowerman:1996}.}
Most people believe that they do this 
as a result of hearing a particular word repeated several times in a
certain context and gradually abstracting an
association between a word and a meaning. But 
how is it that they know which meaning to associate with a 
particular word? How is it that word acquisition goes
so rapid? We will follow a different 
approach, which leaves an open question whether
this applies to human word acquisition as well. The
approach will be selectionist. The agents {\scshape construct} 
hypotheses, either on the basis of one specific case 
where they guess through a non-verbal strategy what 
the meaning of an unknown word might be, 
or they have simply invented a new word because
they do not have one yet. The hypotheses are 
then tried out in subsequent games and either receive confirmation
or are falsified. As a side effect of this local behaviour, 
a global self-organising dynamic process arises leading gradually to coherence.

@@

Here is the simplest possible instance of a
language game, based on the scene in \figref{scene1-1}. 
The speaker, {\bfshape a1}, has picked the triangle as the topic.
To give a verbal hint, he needs 
to conceptualise this topic, which means in this 
specific case, to find a 
category (or set of categories) which distinguishes the 
topic from the other objects in the context. Here 
the context contains only one additional 
object, namely the rectangle. The category
[\textsc{vpos} 0.0–0.5]$_{\mathbf a1}$ (lower) fits the criteria. 
[\textsc{vpos} 0.0–0.5] is valid when \textsc{vpos}$ < 0.5$
which is the case for the triangle but not for the rectangle. 
{\bfshape a1} has an association in his lexicon 
relating [\textsc{vpos} 0.0–0.5]$_{\mathbf  a1}$ with the word \textit{lu}, 
{\bfshape a1} retrieves this word and transmits it to 
the hearer, which is agent {\bfshape a2}.

{\bfshape a2} has stored in his lexicon an
association between \textit{lu} and [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$, and so
hypothesises that [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$ must be the meaning of \textit{lu}. 
When this category is applied to the present scene, in other words,
when {\bfshape a2} filters out the objects whose value for \textsc{vpos} do not 
fall in the region $[0.0–0.5]$, only one 
remaining object is obtained, the triangle. Hence {\bfshape a2} concludes that 
this must be the topic and points to it.

@@

There are (at least) two ways in which a game similar to game 125, 
but played earlier might have failed. First of all it can be that
the speaker does not have a word yet for the meaning he wants
to convey. The speaker then invokes a strategy 
to repair this failure. The simplest strategy is to 
create a new word for the present meaning. This is how 
{\bfshape a1} might have created the word \textit{lu}, and associated it 
with [\textsc{vpos} 0.0–0.5] in his lexicon. Such simple constructive steps cause
new words to enter the lexicon.

Second, it can be that the hearer does not know the word. The 
game then fails and the speaker points to 
the topic so that the hearer can make 
an educated guess what the meaning might have been:
If the hearer is able to recover a possible topic
from the non-verbal hint given by the speaker, 
he himself can seek a distinctive category or set of categories 
that delineates this topic from the other objects in 
the context just as the speaker has done. It is 
possible (although not necessary as we will see)
that the hearer {\bfshape a2} arrives at his 
own version of the same category, namely [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$. The 
hearer now stores in his lexicon a new association between the 
form heard, \textit{lu}, and the guessed meaning [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$. 
Based on this extended lexicon, he will succeed in the same game 
in the future and he can use \textit{lu} himself to verbalise [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$
when talking to other agents. This is the way new words
spread in the population.

@@

The observant reader will have noticed immediately that 
I have swept some important difficulties under the carpet. 
First, it could very well be that unknown to the speaker 
another agent had already used the word \textit{lu} 
for another meaning, so that there are now 
two alternative meanings for \textit{lu} in the lexicon. \textit{lu} 
has become ambiguous. Second, it may be the hearer {\bfshape a2} already 
had a word for [\textsc{vpos} 0.0–0.5]$_{\mathbf  a2}$, for example \textit{bomida}, 
so that there are now two synonyms for [\textsc{vpos} 0.0–0.5] in the lexicon. 
Synonymy and ambiguity are very common in 
natural languages and unavoidable when a group of 
distributed autonomous agents creates a language without
a central co-ordinator. This implies that 
the agents' lexicon must be sophisticated enough 
to support multiple associations. An agent must be able to 
store the different meanings being used for the same word, and the 
different words being used for the same meaning.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap2/figs/competition-1.pdf}}
\caption{\label{competition1}The graph shows the competition between different forms for
expressing one meaning in a population of 20 agents. The graph plots 
the frequency of each form in the population, more precisely the percentage
of agents that prefer a particular word. A complex dynamic
process unfolds with periods where one word (first \textit{xu} and then \textit{fepi}) dominates.} 
\end{figure}
